---
title: "Statistical Learning HW1 Applied"
author: "Yaniv Bronshtein"
date: "10/16/2021"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
(a). *Analysis of Primal Problem*
Give the feasible set, the optimal value, and the optimal solution

*Feasible Set: The interval [2,4]*
$(x-2)(x-4)\le0$
$x-2\le0$ and $x-4\le0$
$x\le2$ and $x\le4$
Thus, the optimal point is x*=2
The optimal value is $2^2+1=5$


(b). *Lagrangian and dual Function*
Plot the objective $x^2+1$ versus x. On the same plot, show the feasible set, optimal optimal point and value, and plot the Lagrangian $L(x,\lambda)$ versus
x for a few positive values of $\lambda$. Verify the lower bound property $p* \ge inf L(x,\lambda)$ for $\lambda\ge0$. 
Derive and sketch the Lagrange dual function g.
```{r}


x <- seq(-5,5, 0.1)
f0 <- x^2+1
f1 <- (x-2)*(x-4) 
par(mfrow=c(1,2))
plot(
  x=x, 
  y=f0,
  main="f0 and f1",
  ylab="",
  xlab="x",
  type="l",
  ylim = c(-5,25),
  col="blue"
)
lines(x=x, y=f1, col="magenta")
legend(
  "topleft",
  c("f0=x^2+1","f1=(x-2)(x-4)"),
  fill=c("blue","magenta")
)

abline(v=2, col='red')
abline(v=4, col='red')
abline(h=0, col='black')
```


```{r}
plot(x=x, 
     y=f0,
     ylab="",
     xlab="x",
     type="l",
     ylim = c(-5,25),
     col="brown",
     main="The lagrangian for various values of lambda"
)
lines(x,f0+1.0*f1, col="green")
lines(x,f0+2.0*f1, col="purple")
lines(x,f0+3.0*f1, col="orange")

legend(
  "topleft",
  c("f0+0.0*f1","f0+1.0*f1", "f0+2.0*f1", "f0+3.0*f1"),
  fill=c("brown","green", "purple", "orange")
)

abline(v=2, col='red')
abline(v=4, col='red')

```
The overlayed plot above demonstrates the Lagrangian with input x and $\lambda$ as
the sum of f_0 and f_1 times a constant $\lambda$
The minimum value of the Lagrangian is always less than p*.
The maximum is reached at a $\lambda$ value of 2 and decreases after that.

```{r}


lambda <- seq(-0.9,16/3,0.1); 
g <- (-9*lambda^2)/(1+lambda) + 1 + 8*lambda
plot(x=lambda, 
     y=g, 
     ylab="g(lambda)", 
     xlab="lambda", 
     type="l", 
     xlim=c(-2,4), 
     ylim=c(-10,6)
)
abline(v=-1, lty='dashed')


```