---
title: "MSDS 534 HW3"
author: "Hema Chandra Yedoti"
date: "10/29/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
rm(list = ls())
```

```{r}
is.acyclic <- function(A){
  ## A is an adjacency matrix
  ## i.e. a_ij = 1 if the edge i->j *is* in the graph
  ##      a_ij = 0 if the edge i->j *is not* in the graph
  if (nrow(A) == 1)
    return(TRUE)
  ## compute transitive closure
  H <- A
  diag(H) <- 1
  repeat {
    HH <- sign(H %*% H)
    if (all(HH == H)) 
      break
    else H <- HH
  }
  diag(H) <- 0
  ## h_ij =1 if there is a directed path from i to j
  l <- H[lower.tri(H)]
  u <- t(H)[lower.tri(t(H))]
  com <- (l & u)
  all(!com)
}


```

```{r, echo = FALSE, warning = FALSE, include = FALSE}
library("gRbase")
library("gRain")
library("gRim")
library("Rgraphviz")
library("glasso")
```

# 2)

```{r, echo = FALSE}
# Storing the gene.txt file as a data frame:

gene <- read.table('../data/gene.txt', header = TRUE)
#df_subset <- df[,c(3,6,4,1)]
#gene = read.table('../data/gene.txt', head = TRUE)
```

### a) 
```{r, echo = FALSE}
# Creating an empty 4x4 adjacency matrix to represent the DAGs on genes GAL1, GAL2, GAL3, GAL7:
A = matrix(c(0,0,0,0,
              0,0,0,0,
              0,0,0,0,
              0,0,0,0),nrow=4,byrow=T)

rownames(A) = c("GAL1", "GAL2", "GAL3", "GAL7")
colnames(A) = c("GAL1", "GAL2", "GAL3", "GAL7")
```

```{r, echo = FALSE}
DAGS = list()

for (a in 1:3) {
  i = 1; j = 2
  
  # reset the previous option we made for this pair
  A[i,j] = 0
  A[j,i] = 0
  
  if (a == 1) {
    A[i,j] = 0
    A[j,i] = 0
  }
  
  if (a == 2) {
    A[i,j] = 1
  }
  
  if (a == 3) {
    A[j,i] = 1
  }
  
  # pair (GAL1, GAL3)
  for (b in 1:3) {
    i = 1; j = 3
    
    # reset the previous option we made for this pair
    A[i,j] = 0
    A[j,i] = 0
    
    if (b == 1) {
      A[i,j] = 0
      A[j,i] = 0
    }
    
    if (b == 2) {
      A[i,j] = 1
    }
    
    if (b == 3) {
      A[j,i] = 1
    }
    
    # pair (GAL1, GAL7)
    for (c in 1:3) {
      i = 1; j = 4
      
      # reset the previous option we made for this pair
      A[i,j] = 0
      A[j,i] = 0    
      
      if (c == 1) {
        A[i,j] = 0
        A[j,i] = 0
      }
    
      if (c == 2) {
        A[i,j] = 1
      }
    
      if (c == 3) {
        A[j,i] = 1
      }
      
      # pair (GAL2, GAL3)
      for (d in 1:3) {
        i = 2; j = 3
        
        # reset the previous option we made for this pair
        A[i,j] = 0
        A[j,i] = 0
        
        if (d == 1) {
          A[i,j] = 0
          A[j,i] = 0
        }
    
        if (d == 2) {
          A[i,j] = 1
        }
    
        if (d == 3) {
          A[j,i] = 1
        }
        
        # pair (GAL2, GAL7)
        for (e in 1:3) {
          i = 2; j = 4
          
          # reset the previous option we made for this pair
          A[i,j] = 0
          A[j,i] = 0
          
          if (e == 1) {
            A[i,j] = 0
            A[j,i] = 0
          }
          
          if (e == 2) {
            A[i,j] = 1
          }
          
          if (e == 3) {
            A[j,i] = 1
          }
          
          # pair (GAL3, GAL7)
          for (f in 1:3) {
            i = 3; j = 4
            
            # reset the previous option we made for this pair
            A[i,j] = 0
            A[j,i] = 0
            
            if (f == 1) {
              A[i,j] = 0
              A[j,i] = 0
              
              # add this matrix to our list if it is acyclic
              if (is.acyclic(A)) {
                DAGS[[length(DAGS) + 1]] = A
              }
            }
            
            if (f == 1) {
              A[i,j] = 1
              
              # add this matrix to our list if it is acyclic
              if (is.acyclic(A)) {
                DAGS[[length(DAGS) + 1]] = A
              }
            }
            
            if (f == 3) {
              A[j,i] = 1
              
              # add this matrix to our list if it is acyclic
              if (is.acyclic(A)) {
                DAGS[[length(DAGS) + 1]] = A
              }
            }
          }
        }
      }
    }
  }
}
```

* There are 6 pairs of nodes and each loop represents one among these.
* We modify i and j in each loop to represent the indices corresponding to those pair of nodes.
* For each pair of nodes, there are 3 options: no edge and inserting an edge with either direction.

```{r, echo = FALSE}
# Function to calculate BIC:
getBIC = function(A, names, df) {
  # Takes an adjacency matrix and column/row names of the nodes and the data as parameters
  # Calculating BIC of the graph by summing BIC for each node by treating each node as the child node:
  dag = as(A, "graphNEL")
  
  bic = 0
  
  for (i in 1:length(names)) {
    # Calculating BIC for i th node:
    child = names[i]
    parents = parents(child, dag)
    
    # if the child has parents, get BIC using linear regression:
    if (length(parents) > 0) {
      model = lm(gene[[child]] ~., data = df[c(child, parents)])
      bic = bic + BIC(model)
    }
    # otherwise, get BIC using formula:
    else {
      N = length(df[,child])
      bic_child = N*log(1/N*sum((df[,child]-mean(df[,child]))^2))
      bic = bic + bic_child
    }
  }
  
  return(bic)
}
```

```{r, echo = FALSE}
# Creating an empty vector to store the BIC of each graph:
BICs = vector(length = length(DAGS))

gene_sub = gene[c("GAL1", "GAL2", "GAL3", "GAL7")]

# Computing BIC for each DAG:
for (i in 1:length(DAGS)) {
  BICs[i] = getBIC(DAGS[[i]], rownames(A), gene_sub)
  }
```

See code in R Markdown file.

### b)
```{r, echo = FALSE}
# Sorting the vector of BICs:
BICs.sorted = sort(BICs)

# 10 Models with least BIC:
BICs.sorted[1:10]
```

* There are 2 models with same BIC (-27452.71).
* So, if we consider the three smallest BIC, we have 4 models.

```{r, echo = FALSE}
smallest = BICs.sorted[1:4]

first_smallest.indices = which(BICs %in% smallest[1:1])
second_smallest.indices = which(BICs %in% smallest[2:2])
third_smallest.indices = which(BICs %in% smallest[3:4])
```

# DAG of the model with the smallest BIC:
```{r, echo = FALSE}
plot(as(DAGS[[first_smallest.indices[1]]], "graphNEL"))

title(main = paste(c("BIC =", smallest[1], "(1st Smallest)"), collapse = " "))
```

# DAG of the model with the 2nd smallest BIC:
```{r, echo = FALSE}
plot(as(DAGS[[second_smallest.indices[1]]], "graphNEL"))

title(main = paste(c("BIC =", smallest[2], "(2nd Smallest)"), collapse = " "))
```

# DAGs of the model(s) with the 3rd smallest BIC:
```{r, echo = FALSE}
par(mfrow = c(1,2))
plot(as(DAGS[[third_smallest.indices[1]]], "graphNEL"))
plot(as(DAGS[[third_smallest.indices[2]]], "graphNEL"))

title(main = paste(c("BIC =", smallest[3], "(3rd Smallest)"), collapse = " "))
```

\newpage

# 3)

### a)

```{r, echo = FALSE}
# Creating a function which implements the step-wise algorithm for estimating the graph structure:

# start with some initial graph and apply the 3 operations to it and choose the one with smallest BIC
# now, start from the chosen graph and apply the 3 operations again and repeat until a smaller BIC is not obtained
# returns a list containing adjacency matrix of this graph and its bic

stepwise = function(df) {
  # Takes a dataframe as input parameter:
  
  # Starting the algorithm with an intitial graph at each iteration
  names = colnames(df)

  # empty graph will be our initial graph for the algorithm
  starting = matrix(0, length(names), length(names))
  rownames(starting) = names
  colnames(starting) = names
  
  min_bic = getBIC(starting, names, df)
  
  smaller_BIC_exists = TRUE
  
  while (smaller_BIC_exists) {
    
    # lists which store BICs and adjacency matrices for all graphs we created in this iteration:
    bic_list = list()
    graph_list = list()
    
    # REMOVING EDGES:
    
    # get indices where matrix has a 1 (this marks edges)
    edge_indices = which(starting == 1, arr.ind = T)

    # make sure the graph has edges we can remove
    if (length(edge_indices) > 0) {
      
      # go through each point in matrix where we found a 1
      for (i in 1:(length(edge_indices)/2)) {
        # change the 1 to be a 0
        temp = starting
        temp[edge_indices[i,1], edge_indices[i,2]] = 0

        # get BIC of this temp graph and store it (only if it is acyclic)
        tempBIC = getBIC(temp, names, df)
        bic_list = c(bic_list, tempBIC)
        graph_list[[length(graph_list)+1]] = temp
      }
    }
    
    
    # ADDING EDGES:
      
    # get indices where matrix has a 0 (this marks where there are no edges)
    noedge_indices = which(starting == 0, arr.ind = T)
    
  
    # make sure the graph has nodes with no edge
    if (length(noedge_indices) > 0) {

      # through each opint in matrix where we found a 0
      for (i in 1:(length(noedge_indices)/2)) {
        
        # make sure this edge is not going to and coming from the same node
        if(noedge_indices[i,1] != noedge_indices[i,2]) {
          temp = starting
          temp[noedge_indices[i,1], noedge_indices[i,2]] = 1

          # get BIC of this temp graph and store it (only if it is still acyclic)
          if (is.acyclic(temp)) {
            tempBIC = getBIC(temp, names, df)
            bic_list = c(bic_list, tempBIC)
            graph_list[[length(graph_list)+1]] = temp
          }
        }
      }
    }
    
    
    # REVERSING EDGES:
    
    # make sure the graph has edges we can reverse
    if (length(edge_indices) > 0) {
      
      for (i in 1:(length(edge_indices)/2)) {
        
        temp = starting
        temp[edge_indices[i,1], edge_indices[i,2]] = 0
        temp[edge_indices[i,2], edge_indices[i,1]] = 1
        
        # get BIC of this temp graph and store it (only if it is still acyclic)
        if (is.acyclic(temp)) {
          tempBIC = getBIC(temp, names, df)
          bic_list = c(bic_list, tempBIC)
          graph_list[[length(graph_list)+1]] = temp
        }
      }
    
    }
    
    # SELECTING GRAPH FOR NEXT ITERATION:
    
    # store min BIC and index where min BIC occurs
    this_min_bic = min(unlist(bic_list))
    this_min_graph = which.min(unlist(bic_list))
    
    # if one of the graphs we generated in this iteration is less than our current min bic
    # update the minimum bic and make the corresponding graph our new starting graph
    if (this_min_bic < min_bic) {
      min_bic = this_min_bic
      starting = graph_list[[which.min(unlist(bic_list))]]
    }
    # we did not find a graph with a smaller BIC so stop the algorithm
    else {
      smaller_BIC_exists = FALSE
    }
  }
  
  graph_and_bic = list(starting, min_bic)
    
  return(graph_and_bic)
}
```

See code in R Markdown file.

### b)

# Trying our program on gene.txt dataset:
```{r, echo = FALSE}
results = stepwise(gene)
```

```{r, echo = FALSE}
stepwise_graph = results[[1]]
stepwise_bic = results[[2]]
```

* With the empty graph as our initial graph, we get the following DAG and BIC using the stepwise algorithm:

```{r, echo = FALSE}
plot(as(stepwise_graph, "graphNEL"))

title(main = paste(c("BIC =", results[2]), collapse = " "))
```

\newpage

# 5)

### a)

```{r, echo = FALSE}
# get the covariance matrix of gene data
S = cov(gene)

# list of rho values we will try
rho_list = seq(0.1, 3, by = 0.2)
```

```{r, echo = FALSE}
# store the adj. matrices of resulting graphs
undirected_graphs = list()

for (i in (1:length(rho_list))) {
  # fit graphical lasso model
  fit = glasso(S, rho = rho_list[i])
  
  # get inverse covariance matrix from model
  theta = fit$wi
  
  # TRANSFORMAING covariance matrix to adjacency matrix to represent undirected graph
  
  # assigning names of genes
  colnames(theta) = names(gene)
  rownames(theta) = names(gene)
  
  # anywhere in the covariance matrix that is not 0, we set to TRUE (1)
  # otherwise, set to FALSE (0)
  adj = (theta != 0)
  
  # TRUE values become 1, FALSE values become 0 which represent the edges
  adj = adj*1

  # set diagonal to 0 so no node has an edge to itself
  diag(adj) = 0
  
  undirected_graphs[[length(undirected_graphs)+1]] = adj
}
```

```{r, echo = FALSE}
par(mfrow = c(2,2))

for (i in (1:4)) {
  plot(as(undirected_graphs[[i]], "graphNEL"))
  title(main = paste(c("Rho =", rho_list[i]), collapse = " "))
}
```

```{r, echo = FALSE}
par(mfrow = c(2,2))

for (i in (5:8)) {
  plot(as(undirected_graphs[[i]], "graphNEL"))
  title(main = paste(c("Rho =", rho_list[i]), collapse = " "))
}
```


```{r, echo = FALSE}
par(mfrow = c(2,2))

for (i in (9:12)) {
  plot(as(undirected_graphs[[i]], "graphNEL"))
  title(main = paste(c("Rho =", rho_list[i]), collapse = " "))
}
```

```{r, echo = FALSE}
par(mfrow = c(2,2))

for (i in (13:length(undirected_graphs))) {
  plot(as(undirected_graphs[[i]], "graphNEL"))
  title(main = paste(c("Rho =", rho_list[i]), collapse = " "))
}
```

* With a smaller value of rho, the graph is more connected. 
* As rho increases, the graph loses its connections and becomes more and more disconnected. 
* When rho > 2.9, the graph is completely disconnected and all of the features are independent.
